---
title: "Exploring SENACE Datastreams"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Exploring SENACE Datastreams}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

This article explains the dataset architecture of the SENACE Open Data Portal and how web portal views map to official REST API datastreams.

```{r setup}
library(senaceR)
```

---

## Web Portal Catalog vs. REST API Datastreams

Visitors to the [SENACE Open Data Portal](https://datosabiertos.senace.gob.pe) will notice **22 dataset items** in the web catalog. However, the official REST API consolidates these into **8 core Datastreams**:

| Web Portal Catalog Category | Web Items | API Datastream | `senaceR` Helper Function |
| :--- | :---: | :--- | :--- |
| Certificación Ambiental | 3 | `CarteraProyectos` | `senace_cartera_proyectos()` |
| Registros Ambientales | 10 | `ConsultorasAmbientales` | `senace_consultoras_ambientales()` |
| Transparencia y Atención | 3 | `SolicitudAcceso`, `Visitas`, `Reclamos` | `senace_solicitudes_acceso()`, `senace_visitas()`, `senace_reclamos()` |
| Planeamiento y Presupuesto | 3 | `GastoEspecifica`, `GastoFuente`, `GastoGenerica` | `senace_gastos_especifica()`, `senace_gastos_fuente()`, `senace_gastos_generica()` |

### Why the difference?

1. **Subdivided Web Views:** On the portal interface, SENACE subdivides datasets for interactive web browsing. For example, *Consultoras Ambientales* is split into 10 separate pages (one per subsector like Mining, Electricity, Agriculture). In contrast, the REST API delivers the **complete master dataset** in a single query (`ConsultorasAmbientales`).
2. **Filtered Project Views:** The portal displays separate cards for Approved, In Evaluation, and Disapproved projects. The API datastream `CarteraProyectos` returns all projects across all evaluation states.
3. **Web-Only Displays:** A few web pages (*Especialistas*, *Participación Ciudadana*, *Viáticos*) are static portal displays without exposed public API endpoints.

---

## Complete Datastream Reference

`senaceR` provides dedicated functions for each of the 8 REST API datastreams:

```{r DatasetsTable, echo = FALSE}
ds <- senace_list_datasets()
knitr::kable(ds, col.names = c("Datastream ID", "Title", "Description"))
```

---

## Accessing Datastreams via API

All helper functions query live endpoints directly using your SENACE API key (`auth_key`). To ensure maximum data accuracy and consistency, `senaceR` focuses strictly on active API datastreams rather than cached or local static snapshots.
