---
title: "Exploring SENACE Datastreams & Offline Mode"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Exploring SENACE Datastreams & Offline Mode}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

This article explains the dataset architecture of the SENACE Open Data Portal, how web portal views map to REST API datastreams, and how `senaceR` manages offline data snapshots.

```{r setup}
library(senaceR)
```

---

## Web Portal Catalog vs. REST API Datastreams

Visitors to the [SENACE Open Data Portal](https://datosabiertos.senace.gob.pe) will notice **22 dataset items** in the web catalog. However, the official REST API consolidates these into **8 core Datastreams**:

| Web Portal Catalog Category | Web Items | API Datastream | `senaceR` Helper Function |
| :--- | :---: | :--- | :--- |
| Certificación Ambiental | 3 | `CarteraProyectos` | `senace_cartera_proyectos()` |
| Registros Ambientales | 10 | `ConsultorasAmbientales` | `senace_consultoras_ambientales()` |
| Transparencia y Atención | 3 | `SolicitudAcceso`, `Visitas`, `Reclamos` | `senace_solicitudes_acceso()`, `senace_visitas()`, `senace_reclamos()` |
| Planeamiento y Presupuesto | 3 | `GastoEspecifica`, `GastoFuente`, `GastoGenerica` | `senace_gastos_especifica()`, `senace_gastos_fuente()`, `senace_gastos_generica()` |

### Why the difference?

1. **Subdivided Web Views:** On the portal interface, SENACE subdivides datasets for interactive web browsing. For example, *Consultoras Ambientales* is split into 10 separate pages (one per subsector like Mining, Electricity, Agriculture). In contrast, the REST API delivers the **complete master dataset** in a single query (`ConsultorasAmbientales`).
2. **Filtered Project Views:** The portal displays separate cards for Approved, In Evaluation, and Disapproved projects. The API datastream `CarteraProyectos` returns all projects across all evaluation states.
3. **Web-Only Displays:** A few web pages (*Especialistas*, *Participación Ciudadana*, *Viáticos*) are static portal displays without exposed public API endpoints.

---

## Complete Datastream Reference

`senaceR` provides dedicated functions for each of the 8 REST API datastreams:

```{r DatasetsTable, echo = FALSE}
ds <- senace_list_datasets()
knitr::kable(ds, col.names = c("Datastream ID", "Title", "Description"))
```

---

## Working Offline & Resilient Data Access

`senaceR` incorporates pre-packaged, compressed snapshots (`.rda`) of all 8 datastreams in `data/`.

### Offline Mode (`offline = TRUE`)

If you are working without an internet connection or require reproducible script execution, pass `offline = TRUE`:

```{r offline-example}
# Retrieve packaged snapshots without network calls
cartera <- senace_cartera_proyectos(offline = TRUE)
dim(cartera)

consultoras <- senace_consultoras_ambientales(offline = TRUE)
dim(consultoras)
```

### Automatic Fallback

When `offline = FALSE` (the default), `senaceR` attempts to fetch live data from the SENACE API. If the API endpoint is unreachable or returns an HTTP error, `senaceR` issues an informative warning and loads the bundled local snapshot so your analysis can continue uninterrupted.


