---
title: "Getting Started with senaceR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with senaceR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

The `senaceR` package provides programmatic access to open data published by SENACE (*Servicio Nacional de Certificación Ambiental para las Inversiones Sostenibles*), the national authority responsible for evaluating detailed environmental impact assessments for major investment projects in Peru.

Through its REST API, SENACE exposes datasets covering environmental certification portfolios, registered consulting firms, institutional visits, public information requests, and budget execution.

`senaceR` simplifies data access by:

1. Handling API authentication (`auth_key`).
2. Formatting JSON API responses into tidy R `tibble` objects.
3. Providing direct, up-to-date access to official REST API datastreams.

---

## Installation & Setup

Install `senaceR` from GitHub:

```r
# install.packages("remotes")
remotes::install_github("PaulESantos/senaceR")
```

Load the package in your session:

```{r setup}
library(senaceR)
```

### Configuring your API Key

To access live data from the SENACE API, register for a free account at [https://datosabiertos.senace.gob.pe](https://datosabiertos.senace.gob.pe) and copy your personal API key (`auth_key`).

You can set your API key for the current R session:

```r
senace_set_key("YOUR_PERSONAL_API_KEY")
```

Or store it permanently in your `.Renviron` file:

```text
SENACE_API_KEY=YOUR_PERSONAL_API_KEY
```

---

## Discovering Datastreams

To view all official datastreams available through the API:

```{r list-datasets}
datasets <- senace_list_datasets()
datasets
```

---

## Accessing Datastreams

`senaceR` provides dedicated helper functions for each datastream:

### 1. Sustainable Investment Project Portfolio

Retrieve the complete status of environmental impact studies evaluated by SENACE:

```r
proyectos <- senace_cartera_proyectos()
```

### 2. Environmental Consulting Firms Register

Retrieve authorized environmental consulting firms across all subsectors (Energy, Mining, Transport, Agriculture, etc.):

```r
consultoras <- senace_consultoras_ambientales()
```

### 3. Public Information Access Requests

Track transparency and public information requests submitted to SENACE:

```r
solicitudes <- senace_solicitudes_acceso()
```

---

## Conclusion

`senaceR` streamlines environmental data science workflows in Peru by bridging SENACE's open data portal directly to R data structures. For more details on datastream architecture, see `vignette("datasets")`.
